%\include{preambleSDL}\include{newcommandsSDL}\include{hyphenationSDL}\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ALL THE ABOVE TO BE COMMENTED OUT FOR COMPLETE DOCUMENT! %%%%%%%%%%%


\chapter*{Acknowledgements}\label{acknowledgements}\hypertarget{acknowledgements}{}
\addcontentsline{toc}{chapter}{Acknowledgements}


A number of individuals played a vital role in enabling the research for and the writing of this grammar. 

\vspace{10pt}
I am particularly indebted to the \PS\ community in Arjeplog for graciously allowing me to document their fascinating language and culture. %I am very pleased to have been allowed to participate in their enthusiasm
%\linebreak
Specifically, I would like to thank Nils-Henrik Bengtsson, Inger and Sven Anders Fjällås, Anders-Erling Fjällås, Elsy Rankvist, and Dagny and Edgar\superS{†} Skaile. 
%\vspace{-10pt}

{\centering\It{Gijtov adnet!}\\}
\vspace{10pt}

My PhD supervisor, Ulrike Mosel, who agreed to the task of guiding me quite late in the project leading up to this book, even after much of the data for the present corpus had already been collected, provided invaluable criticism and played an essential role in the development of this grammar \textendash\ \It{mataa kurusu!} %JW: ‘mataa kurusu’ is Teop for thank you very much, according to Teop lexical database. %, and keeping form and function separate, and meta-language 

My sincere thanks are also due %to: %various other individuals: % in Germany, Sweden, Norway, the Netherlands and the United States: 
to Michael Rießler for his motivation, comments and support, particularly concerning Saami linguistics; 
%Marijn van ’t Veer and Erica Knödler for saving me when my computer died shortly before my final deadline; 
to the Hans Rausing Endangered Languages Project for supporting the \PSDP\ for five years. %, and the Department for Northern European Studies at Humboldt-Universität zu Berlin for hosting that project; 
%and to Bruce Morén-Duolljá and Øystein Vangsnes for allowing me to participate in other Saami linguistics projects and financially survive the final year of my PhD project. 

Many thanks also to:
Yvonne Ericson, 
Ciprian Gerstenberger, 
Martin Hilpert, 
Elena Karvovskaya, 
Erica Knödler, 
Kristina Kotcheva, 
Bruce Morén-Duoll\-já, 
Maar\-ten Mous, 
Stanly Oomen, 
John Peterson, 
Christian Rapold, 
Robert Rißmann, 
Florian Siegl, 
Peter Steggo, 
Mulugeta Tsegaye, 
Øystein Vangsnes, 
Riitta-Liisa Valijärvi and 
Marijn van ’t Veer. 

%all of my friends and colleagues in my unofficial home in Leiden. 

And last but far from least, my appreciation and love go out to Rebecca, Miranda, Reuben, Shirley and Lee for their love and support in the past, present and future!

\vspace{24pt}
\hfill \It{Freiburg, March 2014}










%\end{document}